---
title: "Getting started with IRS normalization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with IRS normalization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
```

IRS estimates sample-specific normalization factors from a set of taxa that is
stable with respect to the variable being tested. The usual workflow has three
steps:

1. fit IRS to the count matrix and sample metadata;
2. inspect the selected references and normalization factors;
3. use the fitted object directly or pass it to a downstream DAA package.

This vignette illustrates the complete workflow with a small simulated data
set. A second vignette, `vignette("vandeputte2017", package = "IRS")`, applies
the same workflow to a published gut microbiome study.

## A small microbiome data set

We simulate 30 taxa measured in 40 samples. The first five taxa have a two-fold
increase in the case group. A sample-specific multiplicative factor changes
the total number of observed counts without changing which taxa are truly
associated with group.

```{r simulate}
set.seed(2026)

n_taxa <- 30
n_samples <- 40
signal_taxa <- paste0("taxon", 1:5)

metadata <- data.frame(
  group = factor(
    rep(c("control", "case"), each = n_samples / 2),
    levels = c("control", "case")
  ),
  row.names = paste0("sample", seq_len(n_samples))
)

sample_scale <- exp(rnorm(n_samples, sd = 0.35))
baseline_mean <- exp(runif(n_taxa, log(15), log(80)))
mean_matrix <- outer(baseline_mean, sample_scale)
mean_matrix[1:5, metadata$group == "case"] <-
  2 * mean_matrix[1:5, metadata$group == "case"]

counts <- matrix(
  rpois(length(mean_matrix), lambda = as.vector(mean_matrix)),
  nrow = n_taxa,
  dimnames = list(
    paste0("taxon", seq_len(n_taxa)),
    rownames(metadata)
  )
)

dim(counts)
table(metadata$group)
```

IRS expects taxa in rows, samples in columns, and matching sample names in the
metadata.

## Fit and inspect IRS

```{r fit}
library(IRS)

fit <- fit_irs(
  counts = counts,
  meta_dat = metadata,
  predictor = "group",
  seed = 123,
  verbose = FALSE
)

fit
```

The fitted object records the final reference set, sample size factors, and
iteration history. Here IRS selects stable taxa while excluding the five taxa
used to generate the group effect.

```{r inspect-fit}
head(reference_taxa(fit))
length(reference_taxa(fit))
intersect(reference_taxa(fit), signal_taxa)

fit_diagnostics <- diagnostics(fit)
fit_diagnostics$converged
fit_diagnostics$iterations
head(fit_diagnostics$samples)
```

Because the sampling factors are known in this example, they can be compared
with the estimated IRS factors.

```{r plot-factors, fig.alt="IRS size factors versus simulated sample scales"}
plot(
  sample_scale,
  size_factors(fit),
  log = "xy",
  pch = 19,
  xlab = "Simulated sample scale",
  ylab = "IRS size factor"
)
abline(lm(log(size_factors(fit)) ~ log(sample_scale)), col = "#2C7FB8")
round(cor(log(sample_scale), log(size_factors(fit))), 3)
```

## Normalize counts and test taxa

`normalized_counts()` divides each sample by its centered IRS reference total
and retains the original taxa-by-samples orientation.

```{r normalize}
irs_counts <- normalized_counts(fit, counts)
irs_counts[1:4, 1:4]
```

For a simple two-group analysis, `irs_wilcox()` combines IRS normalization,
taxon-wise Wilcoxon tests, and multiple-testing adjustment.

```{r wilcoxon}
irs_result <- irs_wilcox(
  fit,
  counts,
  group = "group",
  metadata = metadata
)

irs_hits <- subset(irs_result, p_adjusted < 0.05)
irs_hits[, c("taxon", "statistic", "p_adjusted")]
```

All five simulated signals are recovered in this run.

```{r truth-summary}
data.frame(
  discoveries = nrow(irs_hits),
  true_signals_found = sum(irs_hits$taxon %in% signal_taxa),
  other_taxa_called = sum(!irs_hits$taxon %in% signal_taxa)
)
```

The sample-scale variation makes total-sum scaling a useful contrast in this
example. The same Wilcoxon test after total-sum scaling calls many stable taxa,
whereas IRS retains the five simulated signals.

```{r tss-comparison}
tss <- sweep(counts, 2, colSums(counts), "/")
tss_p <- apply(tss, 1, function(x) {
  wilcox.test(x ~ metadata$group, exact = FALSE)$p.value
})
tss_hits <- rownames(counts)[p.adjust(tss_p, method = "BH") < 0.05]

data.frame(
  normalization = c("Total-sum scaling", "IRS"),
  discoveries = c(length(tss_hits), nrow(irs_hits)),
  true_signals_found = c(
    sum(tss_hits %in% signal_taxa),
    sum(irs_hits$taxon %in% signal_taxa)
  )
)
```

## Continue with a downstream DAA package

The same `fit` can be reused across supported downstream workflows.

| Downstream method | IRS helper | What is passed downstream |
|:--|:--|:--|
| DESeq2 | `inject_deseq2()` | size factors |
| edgeR | `inject_edger()` | effective library sizes |
| MaAsLin2 LM | `prepare_maaslin2()` | normalized feature table |
| ZicoSeq | `zicoseq_irs()` | fixed IRS references and reference totals |
| LDM frequency scale | `prepare_ldm()` | normalized feature table |
| Wilcoxon | `irs_wilcox()` | complete test result |
| DACOMP | `reference_for_dacomp()` | reference-taxon indices |
| ALDEx2 | `denom_for_aldex2()` | denominator-taxon indices |

The following examples are ready to run when the corresponding package is
installed.

### DESeq2

```{r deseq2, eval=FALSE}
dds <- DESeq2::DESeqDataSetFromMatrix(counts, metadata, ~ group)
dds <- inject_deseq2(dds, fit)
dds <- DESeq2::DESeq(dds)
deseq2_result <- DESeq2::results(dds, contrast = c("group", "case", "control"))
```

`inject_deseq2()` preserves the integer count matrix. Once IRS factors have
been injected, proceed directly to `DESeq()`.

### edgeR

```{r edger, eval=FALSE}
design <- model.matrix(~ group, metadata)
y <- edgeR::DGEList(counts)
y <- inject_edger(y, fit)
y <- edgeR::estimateDisp(y, design, robust = TRUE)
ql_fit <- edgeR::glmQLFit(y, design, robust = TRUE)
edger_result <- edgeR::glmQLFTest(ql_fit, coef = "groupcase")
```

The count matrix remains unchanged, and edgeR uses effective library sizes
derived from the IRS reference totals.

### MaAsLin2

```{r maaslin2, eval=FALSE}
m2 <- prepare_maaslin2(fit, counts)

maaslin2_result <- Maaslin2::Maaslin2(
  input_data = m2$input_data,
  input_metadata = metadata,
  output = "maaslin2_irs",
  normalization = m2$normalization,
  analysis_method = m2$analysis_method,
  fixed_effects = "group"
)
```

The helper returns samples in rows and sets MaAsLin2 to use the supplied
normalized values with its linear-model workflow.

### ZicoSeq

```{r zicoseq, eval=FALSE}
zicoseq_result <- zicoseq_irs(
  meta.dat = metadata,
  feature.dat = counts,
  grp.name = "group",
  irs_fit = fit,
  is.winsor = FALSE,
  is.post.sample = TRUE,
  post.sample.no = 25,
  perm.no = 999
)
```

`zicoseq_irs()` runs the ZicoSeq testing workflow with the reference set chosen
by IRS.

### DACOMP and ALDEx2

These packages accept a set of reference taxa and then apply their native
transformations and tests.

```{r reference-adapters, eval=FALSE}
dacomp_refs <- reference_for_dacomp(fit, t(counts))
dacomp_result <- dacomp::dacomp.test(
  X = t(counts),
  y = metadata$group,
  ind_reference_taxa = dacomp_refs,
  test = dacomp::DACOMP.TEST.NAME.WILCOXON
)

aldex_refs <- denom_for_aldex2(fit, counts)
aldex_clr <- ALDEx2::aldex.clr(
  reads = counts,
  conds = metadata$group,
  denom = aldex_refs
)
aldex_result <- ALDEx2::aldex.ttest(aldex_clr)
```

### LDM

```{r ldm, eval=FALSE}
ldm_input <- prepare_ldm(fit, counts)

irs_otu <- as.matrix(ldm_input$otu_table)
ldm_result <- LDM::ldm(
  formula = irs_otu ~ group,
  data = metadata,
  scale.otu.table = ldm_input$scale.otu.table,
  freq.scale.only = ldm_input$freq.scale.only,
  comp.anal = ldm_input$comp.anal,
  n.perm.max = 1000
)
```

The LDM helper prepares IRS-normalized values for the frequency-scale branch.

## A few practical checks

Before fitting a downstream model, it is useful to confirm that:

- count-matrix columns and metadata rows have matching sample names;
- `diagnostics(fit)$converged` is `TRUE` and every reference total is positive;
- downstream software is not asked to estimate a second set of normalization
  factors after IRS values have been supplied.

The companion Vandeputte vignette shows how these steps look with published
data and an external microbial-load measurement.
