---
title: "IRS in the Vandeputte 2017 gut microbiome study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IRS in the Vandeputte 2017 gut microbiome study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
```

Vandeputte et al. measured gut microbial composition and total microbial load
in individuals with Crohn's disease and controls. This makes the study a useful
case study for normalization: IRS can be fitted from the sequencing counts,
while flow-cytometry measurements provide an external quantitative comparison.

The package includes genus-level 16S counts and microbial-load measurements for
95 samples. The data object is a list containing a taxa-by-samples count matrix
and a sample metadata table.

```{r load-data}
library(IRS)
data("vandeputte2017", package = "IRS")

counts <- vandeputte2017$counts
sample_data <- vandeputte2017$sample_data

dim(counts)
table(sample_data$group)
head(sample_data)
```

## Fit IRS

We select references for the Crohn's disease comparison and store the resulting
normalization in a reusable `irs_fit` object.

```{r fit}
fit <- fit_irs(
  counts = counts,
  meta_dat = sample_data,
  predictor = "group",
  seed = 123,
  verbose = FALSE
)

fit
head(reference_taxa(fit), 10)
```

The sample diagnostics summarize library totals, reference totals, and centered
IRS size factors.

```{r diagnostics}
fit_diagnostics <- diagnostics(fit)
fit_diagnostics$converged
fit_diagnostics$iterations
summary(fit_diagnostics$samples$reference_fraction)
```

Library totals and IRS reference totals capture different aspects of each
sample. Their relationship can be inspected before downstream testing.

```{r totals-plot, fig.alt="Sequencing library totals versus IRS reference totals"}
group_color <- ifelse(sample_data$group == "CD", "#D95F02", "#1B9E77")
plot(
  fit_diagnostics$samples$library_total,
  fit_diagnostics$samples$reference_total,
  log = "xy",
  pch = 19,
  col = group_color,
  xlab = "Sequencing library total",
  ylab = "IRS reference total"
)
legend(
  "topleft",
  legend = levels(sample_data$group),
  col = c("#1B9E77", "#D95F02"),
  pch = 19,
  bty = "n"
)
```

## Differential abundance with IRS

The built-in Wilcoxon workflow provides a compact first analysis. We add a
median log2 fold change calculated from the normalized abundances to aid
interpretation.

```{r irs-test}
irs_result <- irs_wilcox(
  fit,
  counts,
  group = "group",
  metadata = sample_data
)

irs_abundance <- normalized_counts(fit, counts)
control_median <- apply(
  irs_abundance[, sample_data$group == "Control", drop = FALSE],
  1,
  median
)
cd_median <- apply(
  irs_abundance[, sample_data$group == "CD", drop = FALSE],
  1,
  median
)
irs_result$median_log2_fold_change <-
  log2((cd_median + 1) / (control_median + 1))

top_irs <- irs_result[order(irs_result$p_adjusted), ]
head(
  top_irs[, c("taxon", "median_log2_fold_change", "p_adjusted")],
  10
)
```

`irs_wilcox()` takes the original count matrix and performs IRS normalization
internally. It is an unadjusted two-group analysis; use a model-based adapter
when the study design requires covariate adjustment, paired or repeated
observations, random effects, or permutation strata.

```{r discovery-count}
irs_called <- irs_result$taxon[irs_result$p_adjusted < 0.05]
length(irs_called)
```

For a count-based model, the same fitted normalization can be injected into a
native DESeq2 workflow.

```{r deseq2, eval=FALSE}
dds <- DESeq2::DESeqDataSetFromMatrix(
  countData = counts,
  colData = sample_data,
  design = ~ group
)
dds <- inject_deseq2(dds, fit)
dds <- DESeq2::DESeq(dds)
deseq2_result <- DESeq2::results(
  dds,
  contrast = c("group", "CD", "Control")
)
```

`inject_deseq2()` preserves the integer counts and supplies IRS size factors.
Continue directly to `DESeq()`, which uses the pre-existing factors. Do not call
`estimateSizeFactors()` afterward or assign a `normalizationFactors(dds)`
matrix, because those operations would replace or override IRS normalization.

Other supported DAA packages use the same `fit`; see
`vignette("downstream-daa", package = "IRS")` for complete adapter examples.

## Comparison with quantitative microbiome profiling

The study's total cell counts can be combined with sequencing relative
abundances to form quantitative microbiome profiles (QMP). Here QMP is used as
an external reference for the Crohn's disease comparison.

```{r qmp}
relative_abundance <- sweep(counts, 2, colSums(counts), "/")
qmp <- sweep(
  relative_abundance,
  2,
  sample_data$microbial_load,
  "*"
)

qmp_p <- apply(qmp, 1, function(x) {
  wilcox.test(x ~ sample_data$group, exact = FALSE)$p.value
})
qmp_q <- p.adjust(qmp_p, method = "BH")
qmp_called <- rownames(qmp)[qmp_q < 0.05]

data.frame(
  comparison = c("IRS-normalized", "QMP", "Shared"),
  taxa = c(
    length(irs_called),
    length(qmp_called),
    length(intersect(irs_called, qmp_called))
  )
)
```

In this analysis, nearly all taxa detected from IRS-normalized sequencing
counts are also detected by the flow-cytometry-based QMP analysis.

## Results across downstream models

The IRS adapters were also evaluated with eight downstream DAA workflows. The
table below reports paired results from the same data and uses the QMP-positive
set above as the external benchmark. `Matched` denotes a discovery in that set;
`other` denotes a discovery outside it.

```{r benchmark-table}
data("irs_daa_benchmark", package = "IRS")
vandeputte_benchmark <- subset(
  irs_daa_benchmark,
  dataset == "vandeputte2017"
)

benchmark_display <- data.frame(
  Method = vandeputte_benchmark$method,
  `Native calls (matched/other)` = sprintf(
    "%d (%d/%d)",
    vandeputte_benchmark$native_discoveries,
    vandeputte_benchmark$native_benchmark_positive,
    vandeputte_benchmark$native_benchmark_negative
  ),
  `IRS calls (matched/other)` = sprintf(
    "%d (%d/%d)",
    vandeputte_benchmark$irs_discoveries,
    vandeputte_benchmark$irs_benchmark_positive,
    vandeputte_benchmark$irs_benchmark_negative
  ),
  `Native F1` = vandeputte_benchmark$native_f1,
  `IRS F1` = vandeputte_benchmark$irs_f1,
  check.names = FALSE
)

knitr::kable(benchmark_display, digits = 3, align = c("l", "c", "c", "r", "r"))
```

Across these paired analyses, the IRS-integrated workflow has higher
concordance with the QMP benchmark for each downstream method. The benchmark
table is precomputed so that the vignette remains quick to build; the data
object records the package versions and analysis settings used.

## Data source

The study is described in Vandeputte et al. (2017), *Quantitative microbiome
profiling links gut community variation to microbial load*, Nature 551,
507--511 ([doi:10.1038/nature24460](https://doi.org/10.1038/nature24460)). The
processed files were obtained from the CellCount_Nishijima_2024 data release
([doi:10.5281/zenodo.14243685](https://doi.org/10.5281/zenodo.14243685)) and are
redistributed under CC BY 4.0.
